---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

# smbr2

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/poissonconsulting/smbr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/poissonconsulting/smbr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/poissonconsulting/smbr/graph/badge.svg)](https://app.codecov.io/gh/poissonconsulting/smbr)
[![License: MIT](https://img.shields.io/badge/License-MIT-green.svg)](https://opensource.org/license/mit/)
<!-- badges: end -->

## Introduction

`smbr2` (pronounced simber two) is an R package to facilitate analyses using [`STAN`](http://mc-stan.org) with [cmdstanr](https://mc-stan.org/cmdstanr/).

`smbr2` was developed as an alternative to [smbr](https://github.com/poissonconsulting/smbr). It uses [cmdstanr](https://mc-stan.org/cmdstanr/) instead of [rstan](https://mc-stan.org/rstan/) as the model-fitting engine. 

The motivation for this is that `cmdstanr` is more up-to-date with latest developments by the Stan dev team (i.e. via [CmdStan](https://mc-stan.org/docs/cmdstan-guide/)). `smbr2` is able to incorporate improved functionality such as the new Stan syntax, faster parallelization, progress communication with parallelization, pedantic Stan model checking, Stan model diagnostics summary, and model fitting with alternative engines such as pathfinder, laplace approximation, varational inference, and optimization. 

Note that some S3 methods for class `smb_model` and `smb_code` are imported from `smbr`. 
`smbr` and `smbr2` are part of the [embr](https://github.com/poissonconsulting/embr) family of packages.
To enable `smbr2` functionality in `embr`, modify the `stan_engine` argument in `embr::analyse()`, e.g. `embr::analyse(stan_engine = 'cmdstan-mcmc')`. 

**The new Stan syntax must be used.** `smbr2` will fail with old Stan syntax. If you don't wish to update an old model, simply leave the `stan_engine` argument as the default, which will use `smbr` instead. 


## Demonstration

```{r, message = FALSE}
library(bauw)
library(ggplot2)
library(magrittr)
library(embr)
library(smbr2)
```

```{r}
# define model in Stan language
model <- model(code = "
  data {
      int nAnnual;
      int nObs;
      array[nObs] int Annual;
      array[nObs] int Pairs;
      array[nObs] real Year;
  }
  parameters {
      vector[nAnnual] bAnnual;
      real log_sAnnual;
      real alpha;
      real beta1;
      real beta2;
      real beta3;
  }
  transformed parameters {
    real sAnnual;
    sAnnual = exp(log_sAnnual);
  }
  model {
      vector[nObs] ePairs;
      log_sAnnual ~ normal(0, 10);
      bAnnual ~ normal(0, sAnnual);
      alpha ~ normal(0, 10);
      beta1 ~ normal(0, 10);
      beta2 ~ normal(0, 10);
      beta3 ~ normal(0, 10);
      for (i in 1:nObs) {
        ePairs[i] = exp(alpha + beta1 * Year[i] + beta2 * Year[i]^2 +
                      beta3 * Year[i]^3 + bAnnual[Annual[i]]);
      }
      target += poisson_lpmf(Pairs | ePairs);
  }
")

# add R code to calculate derived parameters
model %<>% update_model(new_expr = "
  for (i in 1:length(Pairs)) {
    prediction[i] <- exp(alpha + beta1 * Year[i] + beta2 * Year[i]^2 +
                       beta3 * Year[i]^3 + bAnnual[Annual[i]])
  }
")

# define data types and center year
model %<>% update_model(
  select_data = list(
    "Pairs" = integer(), "Year*" = integer(),
    Annual = factor()
  ),
  derived = "sAnnual",
  random_effects = list(bAnnual = "Annual")
)

data <- bauw::peregrine
data$Annual <- factor(data$Year)

set.seed(42)

# analyse
analysis <- analyse(model, data = data, seed = 3L, glance = FALSE, stan_engine = "cmdstan-mcmc")

# analyse pathfinder
analysis_path <- analyse(model, data = data, seed = 3L, glance = FALSE, stan_engine = "cmdstan-pathfinder")

# coefficient table
coef(analysis, simplify = TRUE)

coef(analysis_path, simplify = TRUE)

# trace plots
plot(analysis)
```
```{r, message = FALSE}
# make predictions by varying year with other predictors including the random effect of Annual held constant
year <- predict(analysis, new_data = "Year")
year_path <- predict(analysis_path, new_data = "Year")

years <- dplyr::bind_rows(list("mcmc" = year, "pathfinder" = year_path), .id = "engine")

# plot those predictions
ggplot(data = years, aes(x = Year, y = estimate)) +
  geom_point(data = bauw::peregrine, aes(y = Pairs)) +
  geom_line(aes(color = engine)) +
  geom_line(aes(y = lower, color = engine), linetype = "dotted") +
  geom_line(aes(y = upper, color = engine), linetype = "dotted") +
  expand_limits(y = 0)
```

## Installation

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("poissonconsulting/smbr2")
```

## Citation

```{r, comment="", echo=FALSE}
citation(package = "smbr2")
```

## Contribution

Please report any [issues](https://github.com/poissonconsulting/smbr2/issues).

[Pull requests](https://github.com/poissonconsulting/smbr2/pulls) are always welcome.

## Code of Conduct

Please note that the smbr project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

